Super-resolution of Aeromagnetic Survey Data

### 2023-10-11 manuscript.tex
LS edits from proof read of print copy

This thesis reports on three bodies of work across two topics in geophysics: enhancing the high-frequency content of gridded potential field data using \emph{super-resolution}; and, learning a representative function of a potential field extent using \emph{implicit neural representation}.
The first topic enhances the value of geophysical surveys by predicting high-frequency components in sparsely sampled potential field grids.
The second presents a straightforward neural network framework for high quality grid regularisation and processing.
For each body of work in these topics, real-world geophysical case studies are presented in order to demonstrate the effectiveness and challenges of the method with open access data.
Each body of work extends deep learning methods described in computer science literature the same year as the development of each respective geophysical method, to keep pace with rapid advances in the field of machine learning.

In the topic of super-resolution, the first method uses convolutional neural networks to enhance the resolution of real magnetic textures in state magnetic grids at a fixed four times scale.
This is further explored in the second body of work, where synthetic potential field line data are used for model pretraining, to address the lack of survey data with sufficient line spacing variability.
It quantifies the performance of super-resolution upscaling on real-world low-resolution grids with line spacings between \SI{80}{\m} and \SI{480}{\m}, and investigates the useful limits of upscaling using the method.

The second topic explores a method for learning a representation of a potential field extent from scattered real-world survey data by a contemporary neural network.
The data-driven approach can extract high-resolution grids from the learned representation, as well as calculate the analytical spatial gradients.
Grid rasters and horizontal gradients calculated with the method very closely match their numerical counterparts, however, vertical gradients do not.

Individually, the methods developed in this thesis enhance the information extractable from potential field surveys.
However, in combination, and together they reflect ongoing advances in deep learning and their extension to challenges and opportunities in geophysics.

@@

I wish to thank a great number of people who have contributed to my ongoing development.
First, to my supervisors:
EJ, for bringing together a team of talented and friendly geodata scientists, and caring deeply for them all.
Tom, who set a high bar with his PhD and immediately set about raising his students to the same level.
Daniel, for the meticulous, concise, and actionable reviews and suggestions.
Naveed, whose perspective was invaluable.
Andie, Dale, and Robert, and the rest of my family in NSW --- I didn't get to visit as much as I had originally planned in 2019, but we made the most of it! Thanks for the support and encouragement, and imparting all the passions of rocks, computers, and tinkering that I have today.
Tasman, for being a good friend and inspirational colleague for most of my academic career --- looking forward to sharing a project again sometime, somewhere in the future!
Lu, for sharing your incredible amount of geophysics knowledge, incredible cooking, and love of tennis with me.
Matilda, for being a wonderful friend and the best housemate --- Thanks for being silly with me.
My first floor colleagues and fellow tech geeks, Daniel, David, Chris, Minh, and Peter --- Your friendship and knowledge has been a wonderful environment to grow as a geodata scientist in.
And all my SES, CET, and friends beyond: Alex, Anne, Ben, France, Jackson, Jules, Kailah, Lauren, Lauri, Liam, Lilly, Maddie, Maria, Malvina, Nicolas, Ravi, Sean, Sumail, Vickie, and Ze --- Thank you for years of lunches, dinners, beaches, rock climbing, yoga, and all the opportunities to talk rocks that I've been missing.
And to all the teachers throughout my life, both professional and incidental: Thank you.

@@

\chapter[Magnetic Grid Resolution Enhancement using ML]{Magnetic Grid Resolution Enhancement using Machine Learning}
\label{ch:paper1}
\newrefsection{}
This chapter has been reproduced from the following peer reviewed publication. Formatting has been adjusted for consistency.

>>>>

This thesis reports on three bodies of work across two topics in geophysics: enhancing the high-frequency content of gridded potential field data using \emph{super-resolution}; and, learning a representative function of a potential field extent using \emph{implicit neural representation}.
The first topic enhances the value of geophysical surveys by predicting high-frequency components in sparsely sampled potential field grids.
The second presents a straightforward neural network framework for high quality grid regularisation and processing.
For each body of work in these topics, real-world geophysical case studies are presented in order to demonstrate the effectiveness and challenges of the method with open access data.
Each body of work extends deep learning methods described in computer science literature the same year as the development of the body of work, to keep pace with rapid advances in the field of machine learning.

In the topic of super-resolution, the first method uses convolutional neural networks to enhance the resolution of real magnetic textures in state magnetic grids at four times scale.
This is further explored in the second body of work, where a baseline model is trained with synthetic potential field grid data and fine-tuned with real-world data to address the lack of survey data with sufficient geological representation, and uses a line spacing derived low-resolution transform.
It quantifies the performance of super-resolution upscaling on real-world low-resolution grids with a line spacing from \SI{320}{\m} to \SI{80}{\m}, and investigates the structural accuracy of the result.

The second topic explores a method for learning a representation of a potential field extent from scattered real-world survey data by a contemporary neural network.
The data-driven approach can extract high-resolution grids from the learned representation, as well as calculate analytical spatial gradients.
Grid rasters and horizontal gradients calculated with the method very closely match their numerical counterparts, however, vertical gradients do not.

Individually, the methods developed in this thesis enhance the information extractable from potential field surveys.
However, together they reflect ongoing advances in deep learning and their extension to challenges and opportunities in geophysics.

@@

I wish to thank a great number of people who have contributed to my journey.

First, to my supervisors:
EJ, for bringing together a team of talented and friendly geodata scientists, and caring deeply for them all.
Tom, who set a high bar with his PhD and immediately set about raising his students to the same level.
Daniel, for the meticulous, concise, and actionable reviews and suggestions.
Naveed, whose perspective was invaluable.

My family in NSW, Andie, Dale, and Robert --- I didn't get to visit as much as I had originally planned in 2019, but we made the most of it! Thanks for the support and encouragement, and imparting all the passions of rocks, computers, and tinkering that I have today.

Tasman, for being a good friend and inspirational colleague for most of my academic career --- looking forward to sharing a project again sometime, somewhere in the future!
Lu, for sharing your incredible amount of geophysics knowledge, incredible cooking, and love of tennis with me.
Matilda, for being a wonderful friend and the best housemate --- Thanks for being silly with me.

My first floor colleagues and fellow tech geeks, David, Daniel, Chris, Minh, and Peter --- Your friendship and knowledge has been a wonderful environment to grow in as a geodata scientist.
And all my SES, CET, and friends beyond: Alex, Anne, Ben, France, Jackson, Jules, Kailah, Lauren, Lauri, Liam, Lilly, Maddie, Maria, Malvina, Nicolas, Ravi, Sean, Sumail, Vickie, and Ze --- Thank you for years of lunches, dinners, beaches, rock climbing, yoga, and all the opportunities to talk rocks that I've been missing.

And to all the teachers throughout my life, both professional and incidental: Thank you.

@@

\chapter[Magnetic Grid Resolution Enhancement Using ML]{Magnetic Grid Resolution Enhancement Using Machine Learning}
\label{ch:paper1}
\newrefsection{}
This chapter has been reproduced from the following peer reviewed publication. Formatting has been adjusted for consistency.

### 2023-10-11 paper1.tex
LS edits from proof read of print copy

\begin{abstract}
    Densely sampled geophysical surveys are a key driver for mineral exploration, but sample density, and therefore grid resolution, is limited by survey cost.
    Consequently, computational methods are resorted to for upsampling, or `super-resolving', of gridded geophysical survey data.
    However, existing approaches such as interpolation filters do not leverage high-resolution detail from pre-existing geophysical surveys.
    Through the application of state-of-the-art deep learning super-resolution architectures, accurate high-resolution grids are predicted from corresponding low-resolution grid priors.
    Specifically, the \emph{RDN} and \emph{ESRGAN+} neural network architectures, which were originally developed for enhancing images, are applied to enhance magnetic surveys and trained with high-resolution and low-resolution magnetic grids of the same extent.
    \qty{80}{\metre} cell size magnetic grids are upsampled to \qty{20}{\metre} cell size using this method, and predicted value and structural accuracy comparisons to the corresponding ground truth \qty{20}{\metre} grids are presented over two test sites in the Eastern Goldfields Superterrane, Western Australia.
    The method based on \emph{RDN} achieves \qty{53}{\percent} lower error compared to Bicubic interpolation.
    The case study demonstrates that the deep learning approach can improve the resolution and frequency content of geophysical surveys without requiring additional sampling expense, while remaining accurate against known ground truth surveys.
    Super-resolution may assist in the interpretation of low-resolution survey grids, however these upsampled grids cannot perfectly recreate the accuracy of highly sampled surveys gridded at their optimal cell size.
\end{abstract}

@@

\section{Materials and Methods}
\subsection{Geophysical context}
The Eastern Goldfields Superterrane of Western Australia is extensively covered by aeromagnetic surveys of \qty{400}{\metre} line spacing or better, especially over its South-Western extent.
Large survey extents are government sponsored through the Exploration Incentive Scheme \parencite[e.g.][]{griffinStimulatingGreenfieldsExploration2010}, with these low-resolution regional surveys serving to identify prospective areas to undertake surveys at higher resolution.
A number of these higher resolution surveys have been publicly released to open file.
In the Eastern Goldfields the driver for this extensive magnetic coverage is the prominence of highly susceptible BIF and ultramafic formations, which host several major resource projects including Mt Mason (iron ore), Mt Ida and Comet Vale (gold/silver), and Goongarrie (nickel/copper).
A long geological history provides our test region with a variety of magnetic features and textures to test super-resolution performance with, including structurally complex high intensity formations (BIF and ultramafic units), broad textural regions (granites), and extensive linear features (dykes).
Shear zones and fault systems provide further variation to these structures.
Overall, the selected test region in the Eastern Goldfields Superterrane contains sufficient variety of features on which to examine the effect and value of super-resolution upsampling, and determine the suitability of the method for application in other regions.

@@

\subsection{Accuracy Metrics}
The functional aim of super-resolution geophysics is to accurately predict realistic high frequency details to add to a low-resolution grid.
To quantify this, we calculate root mean square error (RMSE) and structural similarity metrics (SSIM and MS-SSIM), on TMI, high pass filtered TMI, and first vertical derivative (1VD) filtered TMI results for each upsampling method across several sample extents.
Euclidean accuracy measures like RMSE quantify the per-pixel accuracy of a predicted grid compared to its corresponding high-resolution ground truth grid.
For this metric, quality improves as the error value approaches \num{0}.
The Structural Similarity (SSIM) \parencite{wangImageQualityAssessment2004} and Multi-Scale Structural Similarity (MSSSIM) \parencite{wangMultiscaleStructuralSimilarity2003} image quality metrics use sliding windows to quantify image perceptual similarity based on localised structural information.
MS-SSIM extends SSIM by performing a weighted sum of the measure at several different scales.
In both structural metrics, the result is a combination of localised image luminance, contrast, and structural information.
A SSIM value of 1 indicates perfect similarity, and values approaching \num{0} indicate increasing dissimilarity.

@@

\begin{figure}[hbtp]
    \includegraphics[width=\linewidth]{fig/p1/filtered.png}
    \caption[High Pass filtered results]{High pass filtered test tile and its upsampled counterparts.
        The cut-off for the high pass Butterworth filter was selected with wavelength equal to the Nyquist wavelength of the LR grid, i.e.,\ four times the HR Nyquist wavelength.
    }
    \label{fig:filtered}
\end{figure}

@@

\begin{figure}[hbtp]
    \includegraphics[width=\linewidth,trim={0 0 0 0},clip]{fig/p1/jhillvis.jpg}
    \caption[First Vertical Derivative SR TMI over the Mt Mason resource project]{First Vertical Derivative TMI over the Mt Mason resource project, with the super-resolution predictions (d, e) showing a variety of inference artefacts.
    Red asterisks (*) show a vertical tile boundary present as a 3--4 pixel-wide error in predicted TMI value.
    Red caret (\^{ }) indicates a region of subtle ripple-mark textural artefact, only present in ESRGAN+ predictions.
    }
    \label{fig:jhillvis}
\end{figure}

@@

% \section{Declaration of Competing Interest}
% The authors declare the following financial interests / personal relationships which may be considered as potential competing interests: This work was supported by a Rio Tinto Iron Ore PhD scholarship.

>>>>

\bigskip
\begin{abstract}
    Densely sampled geophysical surveys are a key driver for mineral exploration, but sample density, and therefore grid resolution, is limited by survey cost.
    Consequently, computational methods are resorted to for upsampling, or `super-resolving', of gridded geophysical survey data.
    However, existing approaches such as interpolation filters do not leverage high-resolution detail from pre-existing geophysical surveys.
    Through the application of state-of-the-art deep learning super-resolution architectures, accurate high-resolution grids are predicted from corresponding low-resolution grid priors.
    Specifically, the \emph{RDN} and \emph{ESRGAN+} neural network architectures, which were originally developed for enhancing images, are applied to enhance magnetic surveys and trained with high-resolution and low-resolution magnetic grids of the same extent.
    \qty{80}{\metre} cell size magnetic grids are upsampled to \qty{20}{\metre} cell size using this method, and predicted value and structural accuracy comparisons to the corresponding ground truth \qty{20}{\metre} grids are presented over two test sites in the Eastern Goldfields Superterrane, Western Australia.
    The method based on \emph{RDN} achieves \qty{53}{\percent} lower error compared to Bicubic interpolation.
    The case study demonstrates that the deep learning approach can improve the resolution and frequency content of geophysical surveys without requiring additional sampling expense, while remaining accurate against known ground truth surveys.
    Super-resolution may assist in the interpretation of low-resolution survey grids, however these upsampled grids cannot perfectly recreate the accuracy of highly sampled surveys gridded at their optimal cell size.
\end{abstract}

@@

\section{Materials and methods}
\subsection{Geophysical context}
The Eastern Goldfields Superterrane of Western Australia is extensively covered by aeromagnetic surveys of \qty{400}{\metre} line spacing or better, especially over its South-Western extent.
Large survey extents are government sponsored through the Exploration Incentive Scheme \parencite[e.g.][]{griffinStimulatingGreenfieldsExploration2010}, with these low-resolution regional surveys serving to identify prospective areas to undertake surveys at higher resolution.
A number of these higher resolution surveys have been publicly released to open file.
In the Eastern Goldfields the driver for this extensive magnetic coverage is the prominence of highly susceptible BIF and ultramafic formations, which host several major resource projects including Mt Mason (iron ore), Mt Ida and Comet Vale (gold/silver), and Goongarrie (nickel/copper).
A long geological history provides our test region with a variety of magnetic features and textures to test super-resolution performance with, including structurally complex high intensity formations (BIF and ultramafic units), broad textural regions (granites), and extensive linear features (dykes).
Shear zones and fault systems provide further variation to these structures.
Overall, the selected test region in the Eastern Goldfields Superterrane contains sufficient variety of features on which to examine the effect and value of super-resolution upsampling, and determine the suitability of the method for application in other regions.

@@

\subsection{Accuracy metrics}
The functional aim of super-resolution geophysics is to accurately predict realistic high frequency details to add to a low-resolution grid.
To quantify this, we calculate root mean square error (RMSE) and structural similarity metrics (SSIM and MS-SSIM), on TMI, high pass filtered TMI, and first vertical derivative (1VD) filtered TMI results for each upsampling method across several sample extents.
Euclidean accuracy measures like RMSE quantify the per-pixel accuracy of a predicted grid compared to its corresponding high-resolution ground truth grid.
For this metric, quality improves as the error value approaches \num{0}.
The Structural Similarity (SSIM) \parencite{wangImageQualityAssessment2004} and Multi-Scale Structural Similarity (MSSSIM) \parencite{wangMultiscaleStructuralSimilarity2003} image quality metrics use sliding windows to quantify image perceptual similarity based on localised structural information.
MS-SSIM extends SSIM by performing a weighted sum of the measure at several different scales.
In both structural metrics, the result is a combination of localised image luminance, contrast, and structural information.
A SSIM value of 1 indicates perfect similarity, and values approaching \num{0} indicate increasing dissimilarity.

@@

\begin{figure}[hbtp]
    \includegraphics[width=\linewidth,trim={0 0 0 -5mm},clip]{fig/p1/hpfilt.jpg}
    \caption[High Pass filtered results]{High pass filtered test tile and its upsampled counterparts.
        The cut-off for the high pass Butterworth filter was selected with wavelength equal to the Nyquist wavelength of the LR grid, i.e.,\ four times the HR Nyquist wavelength.
    }
    \label{fig:filtered}
\end{figure}

@@

\begin{figure}[hbtp]
    \includegraphics[width=\linewidth,trim={0 0 0 0},clip]{fig/p1/1vd.jpg}
    \caption[First Vertical Derivative SR TMI over the Mt Mason resource project]{First Vertical Derivative TMI over the Mt Mason resource project, with the super-resolution predictions (d, e) showing a variety of inference artefacts.
    Red asterisks (*) show a vertical tile boundary present as a 3--4 pixel-wide error in predicted TMI value.
    Red caret (\^{ }) indicates a region of subtle ripple-mark textural artefact, only present in ESRGAN+ predictions.
    }
    \label{fig:jhillvis}
\end{figure}

@@

% \section{Declaration of competing interest}
% The authors declare the following financial interests / personal relationships which may be considered as potential competing interests: This work was supported by a Rio Tinto Iron Ore PhD scholarship.

### 2023-10-11 paper2.tex
LS edits from proof read of print copy

\section{Introduction}
We propose a method for super-resolution (SR) upscaling of aeromagnetic geophysics data from diverse geological provenance.
This method uses transfer learning \parencite{tanSurveyDeepTransfer2018} from magnetic data forward modelled from a collection of synthetic geological models with varied histories \parencite{jessellNoddyverseMassiveData2022}, and fine tuning with real-world data.
SR employed in this work leverages a contemporary neural network \parencite{leeLocalTextureEstimator2022}, with a realistic method for sampling high-resolution grids and gridding them to low-resolution.
The method is suitable for typical low-resolution regional aeromagnetic surveys in Australia.

@@

\section{Data and Experimental Methods}
\label{sec:2methods}
A realistic geophysical low-resolution transform is developed by simulating aeromagnetic sampling of high-resolution survey grid datasets, and re-gridding the resulting samples.
The proposed dataset methodology uses synthetic forward modelled potential field data to train a selected performant super-resolution network, and the resulting baseline model is subsequently used for transfer learning to fine-tune train the model with real aeromagnetic survey grid data.
The performance of the method is qualitatively described on real survey grid extents, and quantitatively described with standard image quality metrics.

@@

\section{Results}
\label{sec:2results}
\subsubsection{Quantitative Measure}
Due to the dynamic range of magnetic anomaly grids and the relative importance of structural accuracy for gridded data, image quality assessment methods typically used for RGB images such as peak signal-to-noise ratio or mean-squared error are an inadequate quantitative measure of upscaling performance for geophysics.
In addition to mean-square error, FSIM \parencite{linzhangFSIMFeatureSimilarity2011} is used to quantify the quality of the super-resolved image.
FSIM is a full reference image similarity metric that quantifies structural similarity, rather than individual cell values.
%TODO complain about SSIM Structural Similarity (SIM) \parencite{wangImageQualityAssessment2004}
Model performance is quantified using FSIMi in Cref{tab:2metrics}, and it is seen that fine-tune training increases FSIM score for the task of super-resolution for aeromagnetic grids over the baseline model.

>>>>

\section{Introduction}
We propose a method for super-resolution (SR) upscaling of aeromagnetic geophysics data from diverse geological provenance.
This method uses transfer learning \parencite{tanSurveyDeepTransfer2018} from magnetic data forward modelled from a collection of synthetic geological models with varied histories \parencite{jessellNoddyverseMassiveData2022}, and fine tuning with real-world data.
SR employed in this work leverages a contemporary neural network \parencite{leeLocalTextureEstimator2022}, with a realistic method for sampling and gridding high- and low- resolution data.
The method is suitable for typical low-resolution regional aeromagnetic surveys in Australia.

@@

\section{Data and experimental methods}
\label{sec:2methods}
A realistic geophysical low-resolution transform is developed by simulating aeromagnetic sampling of high-resolution survey grid datasets, and re-gridding the resulting samples.
The proposed dataset methodology uses synthetic forward modelled potential field data to train a selected performant super-resolution network, and the resulting baseline model is subsequently used for transfer learning to fine-tune train the model with real aeromagnetic survey grid data.
The performance of the method is qualitatively described on real survey grid extents, and quantitatively described with standard image quality metrics.

@@

\section{Results}
\label{sec:2results}
\subsubsection{Quantitative measure}
Due to the dynamic range of magnetic anomaly grids and the relative importance of structural accuracy for gridded data, image quality assessment methods typically used for RGB images such as peak signal-to-noise ratio or mean-squared error are an inadequate quantitative measure of upscaling performance for geophysics.
In addition to mean-square error, FSIM \parencite{linzhangFSIMFeatureSimilarity2011} is used to quantify the quality of the super-resolved image.
FSIM is a full reference image similarity metric that quantifies structural similarity, rather than individual cell values.
%TODO complain about SSIM Structural Similarity (SIM) \parencite{wangImageQualityAssessment2004}
Model performance is quantified using FSIMi in Cref{tab:2metrics}, and it is seen that fine-tune training increases FSIM score for the task of super-resolution for aeromagnetic grids over the baseline model.
